---
title: "*RegionSetDE*: from peak calls to differential binding"
date: "Last update: `r format(Sys.time(), '%d<sup>th</sup> %B, %Y')`"
author:
- name: "Sebastian Gregoricchio, Ph.D."
  affiliation:
    - "Netherlands Cancer Institute, Oncogenomics division, Amsterdam - Netherlands"
  email: 's.gregoricchio@nki.nl'
package: "*[RegionSetDE](https://github.com/sebastian-gregoricchio/RegionSetDE/)*"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{RegionSetDE: from peak calls to differential binding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = ">",
                      warning = FALSE, message = FALSE, fig.align = "center",
                      dev = "png", dpi = 96, fig.width = 7, fig.height = 4.5)

library(RegionSetDE)
library(dplyr)
library(ggplot2)
```

----------------------------------------------------------

# **Introduction** {#intro}

The [main vignette](RegionSetDE.vignette.html) starts from regions you already have and asks whether the signal over them changes. This one starts a step earlier, from the output of a peak caller, and covers the path that *DiffBind* users will recognise: a sample sheet, a consensus of the peak calls, a count matrix over that consensus, and a differential test.

The two differ in what they take as given. When the regions come from an annotation, they are the same in every sample and the question is only about signal. When they come from peak calls, the regions themselves carry information: a site called in one condition and not in the other is a fact about the experiment, and it is worth keeping next to the statistics rather than discarding it once the consensus is built. Everything below keeps track of which samples and which groups called a peak on each region, and the occupancy stays attached through the counting, the fit and the results table.

<br>

----------------------------------------------------------

# **The example dataset** {#dataset}

Androgen receptor (AR) ChIP-seq in LNCaP prostate cancer cells, from [GSE284522](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE284522) ([Eickhoff *et al.*, *Communications Biology* **8**, 1043, 2025](https://doi.org/10.1038/s42003-025-08449-2)). The cells were hormone deprived and then treated with DMSO, or with the synthetic androgen R1881 for four or twenty-four hours, three biological replicates each, with one input library. Peaks were called with MACS2 in `BAMPE` mode through [SPACCa](https://github.com/sebastian-gregoricchio/SPACCa), on GRCh38.

AR sits in the cytoplasm until it binds a ligand, so this is an experiment where the number of bound sites goes up several fold rather than shifting between locations. That makes it a good illustration and a good trap, for reasons the normalisation section gets to.

The alignments shipped here are a slice: a window of chromosome 19, subsampled to 12 % of the read pairs, with the sequences, the qualities and the read names removed. Counting needs the position and the CIGAR and nothing else, so the counts are the real ones scaled by a known factor and every ratio is preserved. The peaks, on the other hand, are the calls of the complete libraries, filtered to the same window, so they carry the depth the caller actually had. `system.file("extdata", "lncapAR", "SOURCES.md", package = "RegionSetDE")` records exactly what was done, and `inst/scripts/make-example-bams.sh` holds the commands.

<br>

----------------------------------------------------------

# **The sample sheet** {#sheet}

Everything starts from a table with one row per library. `loadSampleSheet()` reads it, recognises the columns whatever they are called, and turns the relative paths into absolute ones, resolved against the folder of the sheet.

```{r load_sheet}
sampleSheet <- loadExampleData("peakSheet")

sampleSheet %>%
  dplyr::select(sample, condition, treatment, time, replicate)
```

The `sample`, `bam`, `peaks` and `input` columns are the ones the package looks for. Names used by *DiffBind* are understood as well, so a sheet written for it can be read without editing: `bamReads` is taken as `bam` and `bamControl` as `input`. Anything else in the table is metadata, kept as it is and available later to the design formula and to the annotations of the plots.

```{r sheet_columns}
colnames(sampleSheet)
```

A sheet carrying only peaks is enough to build a consensus. The BAM files are needed from the counting onwards, and every one of them must be indexed; a BAI or a CSI index both work.

<br>

## Paired-end, single-end, or both

Nothing has to be declared about the layout. `countReads()` takes `pairedEnd = "auto"` by default and reads the flags of the first records of every file, so each library is counted the way it was sequenced. A paired-end fragment is rebuilt from the first mate of each proper pair and counts once; a single-end read is extended to the fragment length of its library, 150 bp unless said otherwise, and counts once as well.

Libraries of both kinds can therefore sit in the same analysis. Each one ends up with one count per sequenced fragment, which is the scale the model needs, and the layout the counting settled on is stored in the `paired.end` column of the sample table so that it can be checked afterwards. The two mistakes this avoids are worth knowing, because they fail in opposite directions: a paired-end file forced through the single-end path counts each mate on its own and almost doubles its values, while a single-end file forced through the paired-end path finds no pair and returns a column of zeros. A test of the package counts one of the libraries below next to a single-end copy of itself, and the two agree to within one per cent.

<br>

## Fragment length of single-end libraries

A single-end read tells where a fragment starts and not how long it is, so the counting has to be told how far to extend it. A single value for every library is rarely right: fragment sizes change with the sonication and the size selection of each library, and a read extended too far spills into the neighbouring regions while one extended too little misses the centre of its own. `fragmentLength` therefore takes one value per sample, in three ways: a vector, a column of the sample sheet, or `"auto"`.

A pipeline that already estimated the fragment length, with phantompeakqualtools for instance, can write it in the sheet, and the column is then named in the counting:

```{r fragment_column, eval = FALSE}
# A 'fragmentLength' column in the sheet, one value per library
counts <- countReads(consensus,
                     sampleSheet = sampleSheet,
                     fragmentLength = "fragmentLength")
```

`"auto"` estimates it from the reads with `estimateFragmentLength()`, which can also be run on its own to look at what it finds. Every fragment leaves a read on the forward strand at its left end and one on the reverse strand at its right end, so the distance between the ends of forward and reverse reads piles up at the fragment length. The libraries here are paired-end and need none of this, but read as single-end, each mate on its own, they show what the estimate does and can be checked against the insert sizes the pairs give directly:

```{r fragment_estimate, fig.height = 5}
fragmentEstimate <- estimateFragmentLength(sampleSheet = sampleSheet,
                                           pairedEnd = FALSE,
                                           verbose = FALSE)

insertSizes <- estimateFragmentLength(sampleSheet = sampleSheet,
                                      verbose = FALSE)

data.frame(sample = fragmentEstimate$table$sample,
           crossCorrelation = fragmentEstimate$table$fragment.length,
           insertSize = insertSizes$table$fragment.length)

fragmentEstimate$plot
```

Five of the six treated libraries land within 15 bp of their median insert size. The DMSO ones come out 27 to 48 bp shorter, and the plot says why: with thirteen peaks in the window there are few fragments piled anywhere, and the cross-correlation is mostly background. The estimate needs enrichment to work on, which is why `"auto"` in `countReads()` reads the regions being counted rather than the whole genome. A second, narrow peak sits at the read length on most libraries; it comes from the parts of the genome where reads cannot be mapped on either strand, and the search starts past it.

The lengths used end up in the `fragment.length` column of the sample table, `NA` for the paired-end samples, and `countBackground()` and `countGreenlist()` reuse them sample by sample.

<br>

----------------------------------------------------------

# **Artefact regions** {#artefacts}

Some stretches of the genome collect reads in every experiment whatever the antibody: satellites, collapsed repeats, the ends of chromosomes, regions whose copy number the cell line has multiplied. A peak called there is real in the sense that the reads are there, and meaningless in every other sense. Two kinds of list take them out.

<br>

## Blacklist

A blacklist is a fixed list per assembly, the same for every experiment. The ENCODE lists for the common assemblies are installed with the package, together with the exclusion set of T2T-CHM13 and the CUT&RUN and CUT&Tag lists of de Mello *et al.*, so no download and no hub are needed.

```{r available_lists}
availableRegionLists(genome = "hg38")
```

```{r blacklist}
blacklist <- loadBlacklist("hg38", seqlevelsStyle = "Ensembl")
blacklist
```

`seqlevelsStyle = "Ensembl"` names the chromosomes the way the peaks and the alignments do. The package would reconcile the two styles on its own when the list is applied, but loading it in the right style from the start keeps what you look at identical to what gets used.

<br>

## Greylist

A greylist is specific to one experiment. It comes from the input library, and flags the places where the input carries far more fragments than the rest of its own genome leads to expect: amplifications of the cell line, repeats the assembly collapsed, regions that stick to any immunoprecipitation. Those are artefacts of this sample rather than of the assembly, so no fixed list catches them. `makeGreylist()` follows GreyListChIP, the procedure *DiffBind* uses: the genome is cut into overlapping windows, a negative binomial is fitted to the counts of the input, and the windows above its 99th percentile are flagged and merged.

```{r greylist_default}
greylist <- makeGreylist(sampleSheet)

S4Vectors::metadata(greylist)$thresholds
```

This is where the example earns its place. The input shipped here is a subsampled slice, about five thousand fragments spread over 12 Mb, so a 1 kb window holds 0.09 fragments on average and the threshold the fit arrives at is **2**. Any window with two fragments is flagged, and about 9 % of the region is thrown out, including several genuine AR sites. The greylist is estimated from the depth of the input, and an input this shallow has no depth to estimate it from. The `thresholds` table is the place to catch it, and the fix is a window wide enough to hold a few fragments each.

```{r greylist}
greylist <- makeGreylist(sampleSheet, binSize = 10000)

S4Vectors::metadata(greylist)$thresholds %>%
  dplyr::select(input, fragments, mean, threshold, regions, greylisted.bp)
```

On a complete input library the default windows are the right choice. What transfers from this example is the habit: look at the threshold before trusting the list.

<br>

## Taking them out

The two lists go in separately, before the consensus is built, through the `blacklist` and `greylist` arguments of `loadConsensusPeaks()` below. Removing the peaks first is better than removing the consensus regions afterwards: an artefact next to a genuine peak would otherwise be merged with it into one region, and the region would then be either kept with the artefact inside or thrown away with the peak.

They are kept apart because they say different things, the blacklist about the assembly and the greylist about the inputs of this experiment, and the object records them the same way `applyBlacklist()` and `applyGreylist()` would: the blacklist in its `blacklist` slot, which follows the regions into the counts and the results and is written down by `exportResults()`, the size of the greylist in its parameters, and the peaks each of them removed from every sample in the filtering log. Once the regions are built, `applyBlacklist()`, `applyWhitelist()` and `applyGreylist()` do the same job on the region set itself.

<br>

----------------------------------------------------------

# **From peak calls to a consensus** {#consensus}

`loadConsensusPeaks()` reads the peak files, removes the peaks in the artefact regions, builds a consensus inside every group, and pools the groups into one region set.

```{r consensus}
consensus <- loadConsensusPeaks(sampleSheet,
                                groupBy = "condition",
                                blacklist = blacklist,
                                greylist = greylist,
                                seqlevelsStyle = "Ensembl")

consensus
```

In this window neither list touches a peak, which is what one hopes; the `n.blacklist` and `n.greylist` columns below say so for every sample, and `filteringLog()` gives the same numbers as a record of the steps, one row per sample and list:

```{r filtering_log}
filteringLog(consensus) %>%
  head(3)
```

`consensusData(consensus)$removed` holds the peaks taken out, with the sample they came from and the list that removed them, empty here. On a whole genome the blacklist alone typically takes a few per cent of the peaks of a transcription factor; the greylist built with the default 1 kb windows [above](#artefacts) would have taken 61 peak calls across the nine samples, genuine AR sites among them, which is the kind of loss this table is there to catch.

`groupBy` names the column that decides which samples have to agree with each other. Choosing it is the one decision that matters here: a consensus built over all nine samples at once would demand agreement between conditions that are not supposed to agree, and a site bound only after treatment would be thrown away before the test ever saw it. Grouping by condition asks for reproducibility between replicates and nothing more, and the union of the groups is what gets counted.

<br>

## The rule a peak has to pass

The consensus inside each group is built by [*consensusRegions*](https://github.com/sebastian-gregoricchio/consensusRegions), and it is not a plain overlap count. Every peak is first placed in one of three classes by its p-value: stringent below `stringencyThreshold` (1e-8), weak up to `weakThreshold` (1e-4), background above it. A peak then needs overlapping peaks in enough of the other replicates, and the evidence of all of them, combined by Stouffer's method, has to clear a threshold. Two consequences follow. A weak peak can be rescued by strong support in the other replicates, which a plain intersection would lose; and a peak present in two replicates out of three can still be dropped when the evidence in both is poor.

How many replicates must hold the peak is `minReplicates`, which counts the replicate the peak came from. The default asks for one supporting replicate, so **two out of three** here. It takes a count, a proportion or a percentage:

```{r min_replicates}
strictConsensus <- loadConsensusPeaks(sampleSheet,
                                      groupBy = "condition",
                                      blacklist = blacklist,
                                      greylist = greylist,
                                      seqlevelsStyle = "Ensembl",
                                      minReplicates = 3,
                                      verbose = FALSE)

data.frame(group = names(consensusData(consensus)$groups),
           twoOfThree = lengths(consensusData(consensus)$groups),
           threeOfThree = lengths(consensusData(strictConsensus)$groups))
```

Asking for all three replicates drops about a quarter of the sites in every group. `minReplicates = 0.66` or `"66%"` write the same thing as a share, which is the more natural way to put it when the groups differ in size. One replicate alone is refused: a peak nobody else confirms is not a consensus. Any other argument of `consensusRegions::buildConsensus()` is passed through the same way, for instance `stringencyThreshold`, `combinationMethod`, `minOverlapFraction` or `mergeMethod`, and `?consensusRegions::buildConsensus` describes them.

The rescue of weak peaks needs weak peaks to work with. The calls shipped here were made at the MACS2 default of *q* < 0.05, and 96 % of them are already stringent, so moving the thresholds changes nothing on this dataset. *consensusRegions* is designed for permissive input, around *p* < 1e-3, and on calls made that way the thresholds matter.

<br>

## Chromosome names

`seqlevelsStyle` is worth setting deliberately. The peak files here name the chromosome `19`, the way Ensembl does, and so do the BAM files, so `"Ensembl"` leaves everything alone. `NULL` does the same by keeping whatever the files hold. Asking for `"UCSC"` renames the peaks and the consensus to `chr19` together.

Mixing the two styles is the classic silent failure of this kind of analysis: an overlap between `chr19` and `19` is not an error, it is simply empty, so a blacklist removes nothing and an occupancy column counts nobody. The package reconciles the styles at every junction instead of trusting them to agree, and stops when it cannot:

| Meeting | What happens |
| :------ | :----------- |
| regions and BAM or bigWig files | the regions are converted to the style of the files for the counting; the object returned keeps its own |
| regions and a blacklist, whitelist or greylist | the list is converted to the style of the regions, and a list sharing no chromosome is refused |
| peaks and the consensus built from them | both follow `seqlevelsStyle` |
| greenlist and BAM files | the greenlist is converted to the style of the files |
| background bins and the counted regions | the bins follow the regions |
| `excludeChromosomes` | read in either style, with a warning when a name reaches no chromosome |

Assemblies are checked separately and more strictly: a list built for another genome is refused rather than overlapped, since rat chr1 and human chr1 share a name and nothing else.

<br>

## What the object holds

```{r consensus_data}
consensusInfo <- consensusData(consensus)

consensusInfo$samples %>%
  dplyr::select(sample, group, n.peaks, n.blacklist, n.greylist)
```

Thirteen peaks without ligand against roughly a hundred after a day of it: the treatment is doing what a nuclear receptor agonist does, and the experiment would be hard to interpret if it were not.

The per-group consensus and the peaks of each sample are kept as well, so nothing has to be recomputed to look at where the groups agree.

```{r consensus_parts}
lengths(consensusInfo$groups)           # the consensus of every group
length(consensusInfo$total)             # the pooled region set that will be counted
```

`consensusGroupList()` returns the consensus of every group as a named list of `GRanges`, the form the functions of *ChIPseeker* take for several peak sets, so the groups can be annotated one by one and compared. `seqlevelsStyle = "UCSC"` renames the chromosomes to match a UCSC `TxDb`, which would otherwise find no gene next to any region named `19`:

```{r group_list, eval = FALSE}
groupList <- consensusGroupList(consensus, seqlevelsStyle = "UCSC")

annotationList <- lapply(groupList,
                         ChIPseeker::annotatePeak,
                         TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene::TxDb.Hsapiens.UCSC.hg38.knownGene)

ChIPseeker::plotAnnoBar(annotationList)
```

Every region carries one logical column per group, plus the number of groups and of samples that called a peak on it.

```{r occupancy_columns}
as.data.frame(S4Vectors::mcols(regionRanges(consensus)$consensus)) %>%
  dplyr::select(dplyr::starts_with("peak.")) %>%
  head(4)
```

`plotPeakUpset()` shows how the groups overlap before any counting happens.

```{r upset}
plotPeakUpset(consensus)
```

Most of the sites are shared by the two R1881 time points, and nearly all of those found without ligand are found with it too. Only one region is specific to DMSO.

<br>

## How similar the peak sets are

The UpSet plot counts regions by group. `plotSampleCorrelation()` with `method = "jaccard"` compares the samples themselves on where their peaks were called, before any read is counted, the occupancy heatmap *DiffBind* draws right after reading the peaks. The Jaccard index of two samples is the number of places where both called a peak over the number where either did, from zero for nothing in common to one for identical peak sets, and the samples are clustered on one minus it.

```{r jaccard, fig.height = 5.5}
plotSampleCorrelation(consensus,
                      method = "jaccard",
                      groupBy = "condition",
                      annotationColumns = c("condition", "replicate"))
```

The places are the regions of the union of all the peaks, so a broad peak and a narrow one on the same site count as a match. `jaccardLevel = "basepair"` counts the bases covered instead, as `bedtools jaccard` does, and then two samples calling the same sites with different widths come out less similar.

The replicates of each condition share between half and three quarters of their peaks, and the two time points of R1881 overlap almost as much, 0.50 to 0.65. DMSO stands apart from both, at 0.10 to 0.19, which is mostly a matter of numbers: thirteen peaks against a hundred cannot overlap much, even when, as here, nearly all of the thirteen are among the hundred. A sample with far fewer peaks than its replicates looks dissimilar to all of them for the same reason, and `consensusData(consensus)$samples` gives the number of peaks to read the index against.

<br>

----------------------------------------------------------

# **Counting** {#counting}

`countReads()` counts fragments over the consensus. Given the sheet it finds the BAM files itself, and takes the sample names and the metadata from the same table.

```{r counting}
counts <- countReads(consensus,
                     sampleSheet = sampleSheet)

counts
```

The sheet names an input for every sample, so the input is counted as well, over the same regions and with the same filters, into an assay of its own beside the counts. Each input file is read once, whatever the number of samples pointing to it, and every sample gets the counts of its own input in its column; a sample without input gets `NA`. Nothing subtracts the input from the counts, which stay the reads of the sample as the model needs them. The input is there to judge the enrichment of the regions, and to check that a change between conditions does not show in the input too, which would point to copy number rather than to binding. `countInput = FALSE` leaves it out.

```{r input_assay}
countTable(counts, input = TRUE, format = "matrix")[1:3, 1:3]
```

Duplicates are dropped and reads below `MAPQ` 20 are ignored, both of which can be changed through `removeDuplicates` and `minMapq`. `excludeChromosomes` leaves chromosomes out of the library sizes, the mitochondrial genome in ATAC-seq above all, and accepts the names in either style. The files are read one after the other here, the slices being small; on whole-genome BAMs `nThreads` spreads them over several workers.

<br>

Every region is counted whole here, one row per region, which is the right choice for a transcription factor whose peaks are a few hundred base pairs wide and change as a block. Counting in tiles, and recentring the regions on the summits as *DiffBind* does, are the two alternatives, and [their own section](#tiles) says when each is worth it.

<br>

## The count table

`countTable()` pulls the values out of any counts, fit or results object, raw or normalised, in three shapes.

```{r count_table_wide}
countTable(counts) %>%
  head(3)
```

The wide format gives the coordinates, the occupancy of every region, and one column per sample. `format = "matrix"` keeps the values only, with the regions as row names, which is what `ComplexHeatmap` or any other tool expects:

```{r count_table_matrix}
countTable(counts, format = "matrix")[1:3, 1:4]
```

`format = "long"` gives one row per region and sample with the sample annotation attached, ready for `ggplot2`:

```{r count_table_long}
countTable(counts, format = "long") %>%
  dplyr::select(region.id, sample, condition, counts) %>%
  head(4)
```

On a tiled object `level` chooses between one row per tile and the tiles summed back into their regions. The same function returns the normalised values, once the normalisation has been run, [below](#normalised_counts).

<br>

## Library summary

`libInfo()` puts the reads of every file next to the ones in the regions.

```{r lib_info}
libInfo(counts, annotationColumns = "condition")
```

The columns measure different things and are not expected to agree. `bam.reads` and `bam.mapped` come from the index of each file, so they cost nothing to read, and they count alignment records: a paired-end fragment is two of them. `library.size` is the number of fragments that went through the filters of the counting, each fragment once, which on paired-end data makes it about half of `bam.mapped`, less the duplicates and the reads of low mapping quality. `reads.in.regions` counts the fragments falling in the consensus, and the FRiP is its share of `library.size`.

Three to four per cent of the fragments fall in the consensus without ligand, against thirteen to twenty-five per cent after it. This is the enrichment the experiment produced, and holding on to it through the normalisation is the whole difficulty of the next section.

The input columns give the same numbers for the input of each sample. The input puts 0.7 % of its fragments in the consensus, which is what the regions collect with no enrichment at all, so even the samples without ligand are about five times enriched over it, and the treated ones eighteen to thirty-five times. A sample whose FRiP sits close to the one of its input carries little signal in the regions, however deep it was sequenced. `fragment.length` is empty here because every library is paired-end.

<br>

## Background bins

`countBackground()` counts the same libraries over wide bins covering everything, and stores the result inside the counts object. It is needed for background normalisation and for the diagnostics that compare one normalisation against another.

```{r background}
counts <- countBackground(counts, binSize = 10000)
```

<br>

----------------------------------------------------------

# **Normalisation** {#normalisation}

This is the section that matters for this dataset.

The methods that come from RNA-seq, TMM, RLE and upper quartile, rest on one assumption: most features do not change, so the middle of the distribution of ratios between two samples is the technical difference between them and can be divided out. Across a transcriptome that usually holds. Here it does not. The ligand brings the receptor to chromatin, the number of bound sites goes up several fold, and most regions really did change in the same direction. Handing that distribution to TMM means telling it that the biology is a loading artefact.

`plotNormComparison()` puts the methods side by side before any of them is committed to.

```{r norm_comparison, fig.height = 5}
plotNormComparison(counts,
                   methods = c("librarySize", "background", "TMM", "readsInRegions"))
```

The consequence is easiest to see by running the same test several times and changing nothing but the normalisation.

```{r norm_effect}
normalisationEffect <-
  lapply(c("librarySize", "background", "TMM", "RLE", "readsInRegions"),
         function(method) {
           normalised <- normalizeCounts(counts, method = method, verbose = FALSE)
           fitted <- fitRegions(normalised, design = ~ condition, engine = "edgeR", verbose = FALSE)
           tested <- resultsTable(testRegions(fitted,
                                              contrast = c("condition", "R1881_24h", "DMSO"),
                                              verbose = FALSE))

           data.frame(method = method,
                      up = sum(tested$diff.status == "up"),
                      down = sum(tested$diff.status == "down"),
                      median.log2FC = round(median(tested$log2FC), 2))
         })

dplyr::bind_rows(normalisationEffect)
```

Same counts, same design, same engine. Library size and background normalisation find the androgen response. TMM and RLE remove most of it. `readsInRegions`, which scales every library by the reads falling inside the consensus, removes it entirely and turns it upside down: that is the scheme that starts by declaring the total signal in the peaks to be equal in every sample, which is precisely the quantity this experiment changed.

The rule this dataset illustrates is worth stating plainly. **If the treatment is expected to change the global amount of binding, the normalisation must not be computed from the bound regions.** What is comparable between the samples is the background, so that is where the factors should come from. When there is no reason to expect a global change, the RNA-seq methods are fine and often better, because they are less sensitive to differences in signal-to-noise between libraries. The point is not that TMM is bad; it is that the assumption behind it is a statement about the experiment, and it should be checked rather than inherited from a default.

<br>

## Greenlist normalisation

This dataset is ChIP-seq and is normalised on its background below. The greenlist is for the case where that is not possible, and the code of this section is shown, not run.

**What a greenlist is.** A set of genomic bins whose background is consistent across CUT&RUN or CUT&Tag experiments, whatever the antibody: regions of steady, low noise, kept at least 5 kb away from genes so that no target is expected to bind there. Since the reads over them do not depend on the target, the differences between samples over them are technical, which makes them a reference built into every library. The lists come from de Mello *et al.* ([*Briefings in Bioinformatics* **25**, bbad538, 2024](https://doi.org/10.1093/bib/bbad538)), who picked the bins by their Shannon entropy across hundreds of public libraries.

**When to use it.** CUT&RUN and CUT&Tag without a spike-in, or with a spike-in that cannot be trusted. Those assays have almost no background, so the wide bins `"background"` normalisation is estimated from hold only a handful of reads, and the region-based methods, TMM or RLE, carry the same assumption that fails in the example above: no global change. The greenlist sits away from the regions of interest and keeps its noise whatever the treatment does to the target, so it holds up when the target changes globally. It is not meant for ChIP-seq, where the background is abundant and does the same job with far more reads behind it.

**How.** Three steps: load the list for the assembly and the assay, count the libraries over it with the same filters as the regions, and normalise.

```{r greenlist, eval = FALSE}
greenlist <- loadGreenlist("hg38", assay = "cuttag")

counts <- countGreenlist(counts, greenlist = greenlist)
counts <- normalizeCounts(counts, method = "greenlist")
```

Lists are installed for hg38 CUT&RUN and CUT&Tag and for mm39 CUT&RUN, and `availableRegionLists(type = "greenlist")` prints them. `countGreenlist()` stores its counts inside the object, where `normalizeCounts()` finds them. The fragments are counted once each, at their centre, which is how the paper counted them with `multiBamSummary --centerReads`. By default the factors are the median of ratios, the size factor DESeq2 computes and the one the greenlist paper used; `greenlistEstimator = "TMM"` or `"sum"` change that. Counts collected with another tool go in through `greenlistCounts`, as a total per sample or a matrix.

`countGreenlist()` holds the libraries to the conditions the lists were built under. The greenlist regions were kept at least 5 kb from genes so that no target binds there, and a greenlist region that one of the regions of the study falls on says the target does bind there in these cells: those are left out, and how many were is reported, unless `excludeCounted = FALSE`. A library below the depth the authors required of a library to enter the construction of the list, 1.5 million reads for human CUT&RUN, 1 million for mouse CUT&RUN and 500,000 for human CUT&Tag, raises a warning, and so does a library reaching fewer than half as many greenlist regions as the median one. `normalizeCounts()` then says how many regions carry reads in every sample, which is what the median of ratios rests on, as it does in DESeq2. The paper sets no rule per group of samples; the number of regions shared by every library is the one that decides how steady the factors are.

What each library holds over the list is kept beside the counts:

```{r greenlist_check, eval = FALSE}
SummarizedExperiment::colData(S4Vectors::metadata(counts)$greenlist)
```

`totals` is the number of fragments on the list, `regions.covered` the regions holding at least one of them, and `library.fraction` their share of the library. The factor of a library is only as steady as the number of fragments behind it. A library holding a handful, which happens on shallow libraries or with a list built for another assay, gets a factor that is mostly noise, and one holding none gets no factor at all.

<br>

## Factors from elsewhere

When the factors come from outside the data, a spike-in counted with another pipeline, deepTools, or a previous analysis, `method = "manual"` takes them as they are. `factorType` says how they apply: `"division"` for size factors the counts are divided by, as DESeq2 writes them, `"multiplication"` for scale factors the counts are multiplied by, as deepTools and most spike-in protocols write them. Named factors are matched to the samples by name and may cover samples absent from the object.

```{r manual_factors}
backgroundCounts <- normalizeCounts(counts, method = "background", verbose = FALSE)

backgroundFactors <- setNames(sampleInfo(backgroundCounts)$scaling.factor,
                              colnames(backgroundCounts))
round(backgroundFactors, 2)

manualCounts <- normalizeCounts(counts,
                                method = "manual",
                                scalingFactors = backgroundFactors,
                                factorType = "division",
                                verbose = FALSE)

all.equal(countTable(manualCounts, normalized = TRUE, format = "matrix"),
          countTable(backgroundCounts, normalized = TRUE, format = "matrix"))
```

The factors of the background normalisation, handed back by hand, give the same normalised values. Getting `factorType` backwards inverts every factor, which on a spike-in experiment turns the largest global change into the smallest; the direction the source writes its factors in is worth checking before anything else.

When the spike-in was aligned separately, `method = "spikeIn"` takes the number of reads each sample put on the exogenous genome, through `spikeInCounts`, and turns them into factors itself.

<br>

## The choice for this dataset

```{r normalise}
counts <- normalizeCounts(counts, method = "background")
```

<br>

## Normalised counts {#normalised_counts}

The factors are stored in the sample table of the object, and the normalised values next to the raw ones. `sampleInfo()` returns the sample table, the annotation of the sheet together with what the package added to it, from a counts object, a fit or the results alike; `columns` keeps only the ones named.

```{r scaling_factors}
sampleInfo(counts, columns = c("sample", "condition", "library.size", "scaling.factor"))
```

`countTable()` returns the normalised values with `normalized = TRUE`, in any of its three formats.

```{r normalised_counts}
countTable(counts, normalized = TRUE) %>%
  head(3)
```

A normalised value is the raw count divided by the scaling factor of its sample, so it stays on the scale of counts rather than of counts per million. `format = "matrix"` gives the values alone, for a heatmap or any other tool, and `format = "long"` puts them in a `norm.counts` column next to the sample annotation. The fit and the results carry the same object, so `countTable(fit, normalized = TRUE)` and `countTable(results$R1881_24h_vs_DMSO, normalized = TRUE)` return the same values later on; the second needs the results to have been built with `carryCounts = TRUE`, which is the default.

What is used for the test is not this table. The model is fitted on the raw counts with the factors as offsets, which keeps the count nature of the data that the negative binomial relies on. The normalised table is for looking at, plotting and exporting.

<br>

----------------------------------------------------------

# **Quality control** {#qc}

Two questions, asked once the normalisation is in place: do the replicates look like each other, and does the structure of the data match the design. They come after the normalisation because an ordination reads distances between samples, and a factor per sample moves every one of those distances; a PCA on raw counts shows the depth of each library as much as its biology. Both functions use the normalised values by default, and `useOffsets = FALSE` switches to the raw ones for comparison.

```{r pca, fig.height = 4.5}
plotRegionPCA(counts, colourBy = "condition", shapeBy = "time")
```

The first component carries most of the variance and separates treated from untreated; the second separates the two durations. Replicates sit together. A component that lines up with the batch instead, or a replicate landing in the wrong group, is worth knowing about before it turns into a dispersion estimate.

`samples` restricts either figure to some of the libraries, by name, by position or with a logical vector, which is how a difference hidden behind the largest one is brought out. The normalisation of the whole analysis is kept, so the values are the ones the model sees.

```{r pca_treated, fig.height = 4.5}
plotRegionPCA(counts,
              samples = sampleInfo(counts)$treatment == "R1881",
              colourBy = "condition")
```

Without DMSO the two durations separate along the first component, which now carries 45 % of the variance, with one replicate of four hours further out than the others.

```{r correlation, fig.height = 5.5}
plotSampleCorrelation(counts,
                      annotationColumns = c("condition", "replicate"),
                      method = "spearman")
```

Any column of the sheet can annotate the heatmap, and the dendrogram is built on one minus the correlation. Rank correlations barely move with the normalisation, since a single factor per sample cannot change the order of the regions within it; the PCA is the one that needs it.

<br>

----------------------------------------------------------

# **Differential binding** {#differential}

The design is an ordinary formula over the columns of the sheet.

```{r fit}
fit <- fitRegions(counts, design = ~ condition, engine = "edgeR")
```

`engine` takes `"edgeR"`, `"DESeq2"` or `"limma"`. The scaling factors computed above enter the fit as offsets, so the model sees the raw counts with the normalisation applied where it belongs rather than a matrix of adjusted values.

<br>

## One contrast

A single comparison is written as a vector of three strings: the column of the sample table, the group of interest, and the group it is compared against.

```{r single_contrast}
resultsEarly <- testRegions(fit,
                            contrast = c("condition", "R1881_4h", "DMSO"),
                            FDR = 0.05,
                            log2FC = 1)

resultsEarly
```

The fold changes read as the second element over the third, so a positive `log2FC` means more signal after four hours of R1881 than in DMSO; swapping the two groups flips every sign and changes nothing else. The groups are written as they appear in the column, and the design does not need to have been set up with either of them as the reference.

<br>

## Every contrast at once

Three conditions give three pairwise comparisons, and they come from the same fit. `pairwiseContrasts()` writes them out from a column of the sample table.

```{r contrasts}
contrasts <- pairwiseContrasts(fit, column = "condition")
contrasts
```

Each level is compared against the ones before it, in the order they appear in the sheet, so DMSO is the reference whenever it is involved and the fold changes read as gains under treatment. `levels` changes the order, and `reference = "DMSO"` would write only the two comparisons against the vehicle. A named list of contrasts runs all of them in one call:

```{r test}
results <- testRegions(fit,
                       contrast = contrasts,
                       FDR = 0.05,
                       log2FC = 1,
                       carryCounts = TRUE)

results
```

`FDR` and `log2FC` set the cut-offs a region has to pass to be labelled `up` or `down`: an adjusted p-value below 0.05 and a fold change beyond two in either direction. They label the regions and do not touch the test; `lfcThreshold` is the argument that moves the fold change inside the test itself. `carryCounts = TRUE` keeps the counts inside the results, which is what lets the plots below draw signal without being handed the counts object again. The multiple testing correction is applied within each contrast, never across them, so every added contrast adds its own share of false positives.

Four hours of androgen already recruit AR to most of its sites, and the second day adds a few more and loses one. The rest of this section uses the twenty-four hour comparison, and shows how to reach the others.

<br>

## The results table

One contrast is taken out of the list by name, and `resultsTable()` returns its table:

```{r results_table}
resultsLate <- results$R1881_24h_vs_DMSO

resultsTable(resultsLate) %>%
  head(3)
```

| Column | Content |
| :----- | :------ |
| `region.set`, `region.id` | the set the region belongs to, and its identifier |
| `seqnames`, `start`, `end`, `width` | its coordinates |
| `log2FC` | the log2 fold change, first level of the contrast over the second |
| `average.signal` | the average abundance of the region over all the samples; for edgeR the log2 counts per million |
| `average.signal.<level>` | the same, over the samples of one condition only |
| `stat` | the test statistic of the engine: the quasi-likelihood F for edgeR |
| `stat.distribution`, `df1`, `df2` | the distribution `stat` follows under the null, and its degrees of freedom |
| `p.value`, `FDR` | the p-value, and its adjustment across all the regions of the contrast |
| `diff.status` | `up`, `down` or `null`, from the `FDR` and `log2FC` cut-offs of `testRegions()` |
| `peak.<group>`, `peak.groups`, `peak.samples` | the occupancy, carried over from the consensus |

The per-condition averages come for every level of the column the contrast compares, all three of them here, so every contrast of the list has the same columns and the stacked table below lines up. `signalBy` names another column of the sample table to average over instead, and `signalBy = FALSE` leaves them out. Read side by side, they say what a fold change is made of: the same log2FC of 3 can be a site going from absent to present, or from bound to more bound, and the averages tell the two apart.

`resultsTable()` on the whole list stacks the contrasts, which is the form to filter with `dplyr`. The `contrast` column in front holds the same names the other functions take, and `contrast.description` says what each one compares:

```{r results_all}
resultsTable(results) %>%
  dplyr::filter(diff.status != "null") %>%
  dplyr::count(contrast, diff.status)
```

`topRegions()` sorts and trims, and takes the contrast by name:

```{r top_regions}
topRegions(results, contrast = "R1881_24h_vs_R1881_4h", n = 5) %>%
  dplyr::select(region.id, log2FC, FDR, peak.R1881_4h, peak.R1881_24h)
```

The strongest change between the two time points is a loss: a site bound after four hours and released by twenty-four, which the comparisons against DMSO alone would not have shown.

<br>

## For a power analysis

The statistic, its distribution and its degrees of freedom are in the table so that the test can be carried further, into a power or sample size calculation with [power4peaks](https://github.com/sebastian-gregoricchio/power4peaks) for instance. `contrastInfo()` gathers the rest of what such a calculation needs, one row per contrast: the engine, the two groups and how many samples each holds, and the degrees of freedom summarised over the regions.

```{r contrast_info}
contrastInfo(results)
```

The statistics themselves are the `stat` column of each contrast, `resultsTable(results$R1881_24h_vs_DMSO)$stat`. The distribution is the one of the engine: an F with `df1` and `df2` degrees of freedom for edgeR, a chi-squared with `df1` when the dispersion was fixed, a t with `df1` for limma, voom and dream, a standard normal for DESeq2. The degrees of freedom of 654 here are not a mistake: on these regions edgeR estimates an infinite prior, every region borrowing all its variability from the others, and caps the total at the residual degrees of freedom of all the regions together, 109 regions times 6.

<br>

## Volcano and MA plots

Every plot takes the contrast by name, and asks for one when the object holds several. The dashed lines mark the cut-offs given to `testRegions()`: the horizontal one the FDR, the two vertical ones the fold change on either side of zero.

```{r volcano}
plotVolcano(results, contrast = "R1881_24h_vs_DMSO", labelTop = 5)
```

```{r volcano_late}
plotVolcano(results, contrast = "R1881_24h_vs_R1881_4h", labelTop = 5)
```

The MA plot puts the fold change against the average abundance, which the volcano hides. It is the place to look for a trend: a cloud leaning up or down with abundance means the normalisation left an intensity-dependent bias behind, and a result sitting on such a trend says more about the normalisation than about the biology.

```{r ma}
plotResultsMA(results, contrast = "R1881_24h_vs_DMSO")
```

The same fold change cut-off appears here as two dashed horizontal lines.

The whole cloud sits above zero, as it should after a ligand that recruits its receptor genome-wide, and it does so at every abundance. The same plot after TMM would have been centred on zero by construction.

```{r heatmap, fig.height = 6}
plotTopHeatmap(results, contrast = "R1881_24h_vs_DMSO", n = 30, annotationColumns = "condition")
```

<br>

## Signal around the regions

A fold change is a single number per region. `plotProfile()` shows the reads behind it, in bins around the centre of every region that changed, as `dba.plotProfile()` does in *DiffBind*: one heatmap per condition, the regions going up and those going down as two groups of rows, and their mean profile on top.

```{r profile, fig.height = 6, fig.width = 8}
plotProfile(results,
            contrast = "R1881_24h_vs_DMSO",
            groupBy = "condition")
```

The signal comes from the BAM files the counts were made from, with the same filters, and is scaled by the factors of the normalisation above, so the panels compare the way the normalised counts do. The rows are sorted by their mean signal and keep the same order in every panel, and the colour scale is shared, so a row read from left to right is one region gaining or losing receptor. Here the 82 regions going up are nearly empty in DMSO and already bound after four hours, with the pile centred on the region; the single region going down is the one site AR leaves.

What the plot guards against is a change spread over the whole window rather than centred on the site, which points to background rather than binding, and a pile sitting away from the centre, which says the regions were not aligned on the signal. The centre here is the midpoint of each consensus region; counted with `summits`, [below](#summits), the regions carry their summit and the profiles are aligned on it. `style = "lines"` draws the mean profiles alone, with their standard error:

```{r profile_lines, fig.height = 3.5}
plotProfile(results,
            contrast = "R1881_24h_vs_DMSO",
            groupBy = "condition",
            style = "lines")
```

The regions need not come from the test. `regions` takes any `GRanges`, a named list of them, one group of rows each, or a region set, and the object only supplies the reads and the normalisation. The consensus split by occupancy, for instance, the sites already bound without ligand against those found only after it:

```{r profile_regions, fig.height = 3.5}
consensusRanges <- regionRanges(consensus)$consensus

plotProfile(counts,
            regions = list(boundInDMSO = consensusRanges[consensusRanges$peak.DMSO],
                           newWithR1881 = consensusRanges[!consensusRanges$peak.DMSO]),
            groupBy = "condition",
            style = "lines")
```

`blacklist` leaves out the rows whose drawn window touches the list, `TRUE` taking the blacklist stored in the object, since a clean region can sit next to an artefact that the window then shows; `whitelist` keeps the regions overlapping a list, promoters for instance; `samples` draws some of the libraries only.

A bigWig file per sample can take the place of the BAM files, through a `bigwig` column of the sample sheet or `signalFiles`; its signal is then drawn as it is, so it has to be normalised already.

<br>

## Occupancy and the test, read together

The point of keeping the peak calls around is to ask whether the two agree. `peakOccupancyTable()` cross-tabulates the result of the test against which groups called a peak.

```{r occupancy_table}
peakOccupancyTable(results, contrast = "R1881_24h_vs_DMSO")
```

```{r occupancy_plot, fig.height = 4.5}
plotPeakOccupancy(results, contrast = "R1881_24h_vs_DMSO")
```

Three kinds of region show up here. Those called in both groups and still changing are sites that were already bound and got more receptor. Those called only after treatment and significant are new sites, the ones the experiment is about. Those called in one group but not changing are the interesting residue: either the caller was near its threshold in the other group, or the difference is real but below the power available in the window. Peak calling is a threshold on a continuum, and a region that fails it is not a region with no signal.

The reverse case deserves its own look. A region called in both conditions and significantly different is invisible to any analysis that works from the peak lists alone, which is the argument for counting rather than intersecting.

<br>

## Writing the results out

```{r export, eval = FALSE}
exportResults(results,
              path = "results",
              onlyChanging = TRUE,
              splitByDirection = TRUE)
```

One table and one BED file per contrast, the BED coloured by direction so it can be dropped into a genome browser as it is, and a record of every parameter the analysis ran with.

<br>

----------------------------------------------------------

# **Whole regions, tiles or summits** {#tiles}

The analysis above counts every consensus region whole. That is the classic choice and the right one for this dataset, but it is not the only way to turn a set of peaks into rows of a count matrix, and the other two are worth knowing about.

<br>

## Whole regions

One row per region, one number per sample. The test asks whether the region as a whole gained or lost signal. This suits transcription factors and ATAC-seq, whose peaks are a few hundred base pairs wide and change as a block, and *DiffBind* did it this way before version 3.0. Its weak point is the width of the regions: a consensus merges overlapping peaks, so the regions come out uneven, from 242 bp to 1.3 kb in this dataset, and a wide region collects more background around the same site than a narrow one.

<br>

## Tiles

`tileWidth` cuts every region into tiles of that width, and each tile becomes a row of its own. The code of this part is shown, not run.

```{r tiles_count, eval = FALSE}
tiledCounts <- countReads(consensus,
                          sampleSheet = sampleSheet,
                          tileWidth = 200)
```

From there the steps are the usual ones, background, normalisation and fit, and `testRegions()` recognises the tiled object. Every tile is tested on its own and the p-values are then combined back into one per region, so the results keep one row per region and the region stays the unit of inference.

```{r tiles_test, eval = FALSE}
tiledCounts <- countBackground(tiledCounts, binSize = 10000)
tiledCounts <- normalizeCounts(tiledCounts, method = "background")

tiledFit <- fitRegions(tiledCounts, design = ~ condition, engine = "edgeR")

tiledResults <- testRegions(tiledFit,
                            contrast = c("condition", "R1881_24h", "DMSO"),
                            combineMethod = "simes")

resultsTable(tiledResults)     # one row per region
tileTable(tiledResults)        # one row per tile
```

Tiles are worth it when the shape of a region matters and not only its total:

* **Broad domains.** H3K27me3, H3K36me3 or H3K9me3 cover tens of kilobases, and a change can be confined to one end. Counted whole, a domain that gains signal on one flank and loses it on the other can come out flat.
* **Partial changes.** A super-enhancer where one constituent gains binding, or a promoter where the signal moves downstream into the gene body.
* **Regions of very different width.** Tiles give every row the same width, so a 20 kb domain and a 500 bp peak are tested on the same scale.

How the tiles of a region are put together is a choice about the question. `combineMethod = "simes"`, the default, asks whether any part of the region changed, and one strong tile is enough. `"holm-min"` asks several tiles to change together, three or 40 % of the region, whichever is more, which suits broad domains where a single tile moving on its own is more often noise than biology. The results table then reports `n.tiles`, and `n.tiles.up` and `n.tiles.down`, the tiles changing significantly in each direction within the region. The `log2FC` of a tiled region is the one of the tile carrying the p-value, not an average, since that is the quantity the p-value is about.

For a transcription factor like AR, tiles add rows without adding information: a peak of 400 bp cut into 200 bp tiles gives two tiles that nearly always move together, and a test with more rows and the same answer.

<br>

## Recentring on the summits {#summits}

The other answer to uneven widths goes the opposite way: instead of cutting the regions, rebuild them at a fixed width around the summit of the signal. This is what *DiffBind* has done by default since version 3.0, with windows of 401 bp, and `summits` in `countReads()` does the same: every region becomes a window of `2 * summits + 1` bp centred on its summit.

```{r summits}
summitCounts <- countReads(consensus,
                           sampleSheet = sampleSheet,
                           summits = 200,
                           verbose = FALSE)

head(SummarizedExperiment::rowRanges(summitCounts)[, c("region.id", "summit")], 3)

table(BiocGenerics::width(SummarizedExperiment::rowRanges(summitCounts)))
```

The summit is found in the reads, as *DiffBind* finds it: the highest point of the fragment pileup of each sample, averaged over the samples with weights proportional to the height of their pileup over their depth, so that the samples carrying the signal decide where it sits. A region with no fragment keeps its midpoint. `region.id` still holds the coordinates of the consensus region each window came from, and `summit` where it was centred. The inputs are counted over the same windows.

`summitSource = "peaks"` takes the summits the peak caller wrote in the tenth column of the narrowPeak files instead, averaged over the peaks overlapping the region with weights proportional to their significance. It needs no BAM file, and on this dataset the two place the summits a median of 26 bp apart. `summits = 0` locates the summits and leaves the regions alone, which keeps the counts of whole regions while letting `plotProfile()` align them on the signal.

That puts every site on the same footing and centres the count on where the protein sits, which is what a transcription factor or ATAC-seq analysis usually wants. Widths between 200 and 500 bp suit most factors. It is wrong for broad marks, whose summit is an arbitrary point in a domain: collapsing the domain to a point throws away the thing being measured, and tiles are the tool there.

On this dataset the summits sit a median of 32 bp from the midpoints of the consensus regions, and 236 bp at most, and none of the windows overlap. The test comes out the same as on whole regions, 82 regions up and one down against DMSO, which is what one expects from a factor whose peaks are narrow and centred already; on a consensus of wide merged regions the two can differ more.

`consensusRegions` offers a recentring of its own, `recentre = TRUE`, which `loadConsensusPeaks()` passes through. It works inside each group, on the summit of the most significant peak of every consensus region, before the groups are pooled; when the summits of two groups sit a few base pairs apart their windows overlap and are merged back, so the pooled regions are close to the width asked for rather than equal to it, and the package says so with a warning. `summits` in `countReads()` works on the pooled regions and gives them all the same width.

<br>

----------------------------------------------------------

# **Where to go next** {#next}

The consensus built here is an ordinary region set, so everything in the main vignette applies to it unchanged. `loadConsensusPeaks()` takes a `regionSets` argument that splits the consensus by overlap with an annotation, so the peaks can be divided into promoters, enhancers and the rest at the moment they are loaded. The set then becomes the unit of the test through `testRegionSets()`, which asks whether a whole class of elements moved instead of counting how many of its members passed a threshold.

<br>

----------------------------------------------------------

# **Session info** {#sessioninfo}

```{r session_info}
sessionInfo()
```
